---
title: "Adding concept intersections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a03_addingConceptIntersections}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Concept sets play an import role when working with data in the format of the OMOP CDM. They can be used to create cohorts after which, as we've seen in the previous vignette, we can identifying intersections between these cohorts. However, PatientProfiles adds another option for working with concept sets which is use them directly for adding associated variables.

It is important to note, and is explained more below, that results may differ when generating a cohort and then identifying intersections between two cohorts compared to working directly with concept sets. The creation of cohorts will involve the collapsing of overlapping records as well as imposing certain requirements such as only including records that were observed during an individuals observation period. When adding variables based on concept sets we will be working directly with record-level data in the OMOP CDM clinical tables.

## Adding concept based variables

For this vignette we'll use the Eunomia synthetic dataset. First lets create our cohort of interest, individuals with an ankle sprain.

```{r}
library(CDMConnector)
library(CodelistGenerator)
library(PatientProfiles)

con <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = CDMConnector::eunomia_dir())
cdm <- CDMConnector::cdm_from_con(con,
                                   cdm_schem = "main",
                                   write_schema = "main")

cdm <- generateConceptCohortSet(
  cdm = cdm,
  name = "ankle_sprain",
  conceptSet = list("ankle_sprain" = 81151),
  end = "event_end_date",
  limit = "all",
  overwrite = TRUE
)

cdm$ankle_sprain
```

Now let's say we're interested in summarising use of acetaminophen among our ankle sprain cohort. We can start by identifying the relevant concepts.

```{r}
acetaminophen_cs <- getDrugIngredientCodes(cdm = cdm, 
                        name = c("acetaminophen"))

acetaminophen_cs
```

Once we have our codes for acetaminophen we can create variables based on these. As with cohort intersections, PatientProfiles provides four types of functions for concept intersections. 

First, we can add a binary flag variable indicating whether an indivdidual had a record of acetaminophen on the day of their ankle sprain or up to 30 days afterwards.  
```{r}
cdm$ankle_sprain %>%
  addConceptIntersectFlag(conceptSet = acetaminophen_cs, 
                          indexDate = "cohort_start_date", 
                          window = c(0, 30)) %>% 
  dplyr::glimpse()
```

Second, we can count the number of records of acetaminophen in this same window for each individual.
```{r}
cdm$ankle_sprain %>%
  addConceptIntersectCount(conceptSet = acetaminophen_cs, 
                          indexDate = "cohort_start_date", 
                          window = c(0, 30)) %>% 
  dplyr::glimpse()
```

Third, we could identify the first start date of acetaminophen in this window.
```{r}
cdm$ankle_sprain %>%
  addConceptIntersectDate(conceptSet = acetaminophen_cs, 
                          indexDate = "cohort_start_date", 
                          window = c(0, 30), 
                          order = "first") %>% 
  dplyr::glimpse()
```

Or fourth, we can get the number of days to the start date of acetaminophen in the window.
```{r}
cdm$ankle_sprain %>%
  addConceptIntersectDays(conceptSet = acetaminophen_cs, 
                          indexDate = "cohort_start_date", 
                          window = c(0, 30), 
                          order = "first") %>% 
  dplyr::glimpse()
```

## Adding multiple concept based variables

We can add more than one variable at a time when using these functions. For example, we might want to add variables for multiple time windows. 

```{r}
cdm$ankle_sprain %>%
  addConceptIntersectFlag(conceptSet = acetaminophen_cs, 
                          indexDate = "cohort_start_date", 
                          window = list(c(-Inf, -1),
                                        c(0, 0),
                                        c(1, Inf))) %>% 
  dplyr::glimpse()
```

Or we might want to get variables for multiple drug ingredients of interest. 

```{r}
meds_cs <- getDrugIngredientCodes(cdm = cdm, 
                                  name = c("acetaminophen",
                                           "amoxicillin",
                                           "aspirin",
                                           "heparin",
                                           "morphine",
                                           "oxycodone",
                                           "warfarin"))

cdm$ankle_sprain %>%
  addConceptIntersectFlag(conceptSet = meds_cs, 
                          indexDate = "cohort_start_date", 
                          window = list(c(-Inf, -1),
                                        c(0, 0))) %>% 
  dplyr::glimpse()
```

## Cohort-based versus concept-based intersections

TO DO

