---
title: "Summarise patient characteristics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a06_summarisePatientCharacteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PatientProfiles)
```


## Characteristics plots

`plotCharacteristics` can plot the summariseCharacteristics results, with boxplot or barplot. First we create a mock data.

```{r, eval=TRUE}
observation_period <- dplyr::tibble(
  observation_period_id = c(1, 2, 3),
  person_id = c(1, 2, 3),
  observation_period_start_date = as.Date(c(
    "1985-01-01", "1989-04-29", "1974-12-03"
  )),
  observation_period_end_date = as.Date(c(
    "2011-03-04", "2022-03-14", "2023-07-10"
  )),
  period_type_concept_id = 0
)
dus_cohort <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 1, 2),
  subject_id = c(1, 1, 2, 3),
  cohort_start_date = as.Date(c(
    "1990-04-19", "1991-04-19", "2010-11-14", "2000-05-25"
  )),
  cohort_end_date = as.Date(c(
    "1990-04-19", "1991-04-19", "2010-11-14", "2000-05-25"
  ))
)
comorbidities <- dplyr::tibble(
  cohort_definition_id = c(1, 2, 2, 1),
  subject_id = c(1, 1, 3, 3),
  cohort_start_date = as.Date(c(
    "1990-01-01", "1990-06-01", "2000-01-01", "2000-06-01"
  )),
  cohort_end_date = as.Date(c(
    "1990-01-01", "1990-06-01", "2000-01-01", "2000-06-01"
  ))
)
medication <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 2, 1),
  subject_id = c(1, 1, 2, 3),
  cohort_start_date = as.Date(c(
    "1990-02-01", "1990-08-01", "2009-01-01", "1995-06-01"
  )),
  cohort_end_date = as.Date(c(
    "1990-02-01", "1990-08-01", "2009-01-01", "1995-06-01"
  ))
)
emptyCohort <- dplyr::tibble(
  cohort_definition_id = numeric(),
  subject_id = numeric(),
  cohort_start_date = as.Date(character()),
  cohort_end_date = as.Date(character())
)
cdm <- mockPatientProfiles(
  dus_cohort = dus_cohort, cohort1 = emptyCohort,
  cohort2 = emptyCohort, observation_period = observation_period,
  comorbidities = comorbidities, medication = medication
)

cdm$dus_cohort <- omopgenerics::newCohortTable(
  table = cdm$dus_cohort, cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2), cohort_name = c("exposed", "unexposed")
  )
)
cdm$comorbidities <- omopgenerics::newCohortTable(
  table = cdm$comorbidities, cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2), cohort_name = c("covid", "headache")
  )
)
cdm$medication <- omopgenerics::newCohortTable(
  table = cdm$medication,
  cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2, 3),
    cohort_name = c("acetaminophen", "ibuprophen", "naloxone")
  ),
  cohortAttritionRef = NULL
)
characteristicsResult <- summariseCharacteristics(
  cdm$dus_cohort,
  cohortIntersect = list(
    "Medications" = list(
      targetCohortTable = "medication", value = "flag", window = c(-365, 0)
    ), 
    "Comorbidities" = list(
      targetCohortTable = "comorbidities", value = "flag", window = c(-Inf, 0)
    )
  )
)
```

Now we show barplot example by setting plotStyle = "barplot" in `plotCharacteristics`. Similar to previous function, user can define axes, facetVars and colorVars. Currently, barplot is only plotting percentage.

```{r, eval=TRUE}
plotCharacteristics(
    data =  characteristicsResult,
    xAxis = "estimate_value",
    yAxis = "variable_name",
    plotStyle = "barplot",
    facetVars = c("group_level"),
    colorVars = c("variable_name", "variable_level")
  )

```

User can plot boxplot based on q25 q75 median min max in data, using `plotCharacteristics` and set plotStyle = "boxplot". Boxplot will be horizontal if the xAxis is set to  estimate_value, and vertical if yAxis is set to estimate_value. But for all the plots, at least one of the xAxis or yAxis has to be estimate_value.

```{r,eval=TRUE}
plotCharacteristics(
    data =  characteristicsResult,
    xAxis = "estimate_value",
    yAxis = "variable_name",
    plotStyle = "boxplot",
    facetVars = "variable_name",
    colorVars = c("group_level")
  )
```
