---
title: "Summarise cohort overlap"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{a08_summariseCohortOverlap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(CDMConnector)
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = tempdir())
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!eunomia_is_available()) downloadEunomiaData()
```

When we create a set of cohorts a common question is what proportion of individuals in one cohort are also in another. The function `summariseCohortOverlap()` can help us to answer such a question. To show how this works lets first create a set of medication cohorts.

```{r, warning = FALSE, message = FALSE}
library(CDMConnector)
library(CodelistGenerator)
library(PatientProfiles)
library(dplyr)
library(ggplot2)

con <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = CDMConnector::eunomia_dir())
cdm <- CDMConnector::cdm_from_con(con,
                                  cdm_schem = "main",
                                  write_schema = "main", 
                                  cdm_name = "Eunomia")

meds_cs <- getDrugIngredientCodes(cdm = cdm, 
                                  name = c("acetaminophen",
                                           "amoxicillin",
                                           "aspirin",
                                           "heparin",
                                           "morphine",
                                           "oxycodone",
                                           "warfarin"))

cdm <- generateConceptCohortSet(
  cdm = cdm,
  name = "meds",
  conceptSet = meds_cs,
  end = "event_end_date",
  limit = "all",
  overwrite = TRUE
)
```

Once we have created our set of cohorts, we can now quickly get a summary of cohort overlaps.

```{r}
meds_overlap <- cdm$meds |> 
  summariseCohortOverlap()

meds_overlap |> 
  glimpse()
```

After this we can create a table with our results.

```{r}
tableCohortOverlap(meds_overlap)
```

Or we could also visualise these results in a plot.

```{r, fig.width=7, fig.height=6.5, message=FALSE, warning=FALSE}
plotCohortOverlap(meds_overlap) +
  theme(legend.position = "top")
```
