---
title: "Get cohort intersections"
author: "Martí Català, Mike Du, Yuchen Guo, Kim López-Güell, Xintong Li, Núria Mercadé-Besora, and Edward Burn"
date: "`r Sys.Date()`"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Get cohort intersections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

In this vignette we present how functions from this package can be used to get intersections between cohorts. This can be useful, for instance, if we want to identify patients with previous conditions. 

The PatientProfiles package is designed to work with data in the OMOP CDM format, so our first step is to create a reference to the data using the DBI and CDMConnector packages. The connection to a Postgres database would look like:

```{r, eval=FALSE}
library(DBI)
library(CDMConnector)

# The input arguments provided are for illustrative purposes only and do not provide access to any database.

con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = "omop_cdm",          
                      host = "10.80.192.00",
                      user = "user_name",
                      password = "user_pasword"
)

cdm <- CDMConnector::cdm_from_con(con,
                                  cdm_schema = "main",
                                  write_schema = "main",
                                  cohort_tables = "cohort_example"
)
```

In this vignette we will work with simulated data generated by the `mockPatientProfiles()` function provided in this package, which mimics a database formatted in OMOP. 

```{r, message= FALSE, warning=FALSE}
library(PatientProfiles)
library(duckdb)
library(tibble)
library(dplyr)

cdm <- mockPatientProfiles(
  patient_size = 1000, 
  drug_exposure_size = 1000
)
                          
```

In this mock dataset there are the following cohort tables:
```{r, message= FALSE, warning=FALSE}
cdm$cohort1 %>%
  glimpse()

cdm$cohort2 %>%
  glimpse()                       
```
## Example: addCohortIntersectFlag and addCohortIntersectCount functions
`addCohortIntersectFlag()`: adds a binary column indicate intersection with a cohort in a time frame.

Suppose cohort2 with definition_id = 1 contains stroke occurrences. If we want to exclude patients from cohort1 who had a stroke event in the last 180 days before entering the cohort, we can use the `addCohortIntersectFlag()` like this:

```{r, message= FALSE, warning=FALSE}
cdm$cohort1WashOut <- cdm$cohort1 %>%
  addCohortIntersectFlag(
    cdm = cdm,
    targetCohortTable = "cohort2",
    window = list(c(-180,-1)),
    targetCohortId = 1,
  ) %>%
  filter(cohort_1_m180_to_m1 == 0)

cdm$cohort1WashOut %>%
  glimpse()
```

`addCohortIntersectCount()`: adds a column where it indicates the number of intersections in a certain time window.

We can use the function to count the number of occurrences of an event of interest in different time windows before entering the study population. For example, we can look at the number of strokes in the 0-90 day, 90-365 day, and all prior history windows:

```{r, message= FALSE, warning=FALSE}
cdm$cohort1StrokeCounts <- cdm$cohort1 %>%
  addCohortIntersectCount(cdm = cdm,
    targetCohortTable = "cohort2",
    window = list(c(-Inf,-366), c(-365,-181), c(-180,-1)),
    targetCohortId = 1
  )

cdm$cohort1StrokeCounts %>%
  glimpse()
```


## Example: addCohortIntersectTime function

`addCohortIntersectTime()`: adds a new column that indicates the number of days in which the subject intersects with another cohort during a specific time frame. If there are multiple intersections, only one will be computed, either the first or the latest one in the time window ("order" argument).

The function can be utilized to calculate the time to the event of interest, such as time until the first stroke after index date. If the patient did not experience the event, the function will return NA.

```{r, message= FALSE, warning=FALSE}
cdm$cohort1TimeTo <- cdm$cohort1 %>%
  addCohortIntersectTime(cdm = cdm,
    targetCohortTable = "cohort2",
    targetCohortId = 1,
    order = "first",
  )

cdm$cohort1TimeTo %>%
  glimpse()
```
## Example: addCohortIntersectDate function

`addCohortIntersectDate()`: appends a column containing the start date of cohorts that are present in a certain window.

This function can be handy in obtaining the date of the next occurrence of a specific event. For instance, suppose cohort1 comprises patients who enrolled when they received their first vaccine dose. We could use this function to obtain the date of their second dose if we have a cohort with vaccine records (e.g. cohort2): 

```{r, message= FALSE, warning=FALSE}
cdm$cohort1NextEvent <- cdm$cohort1 %>%
  addCohortIntersectDate(cdm = cdm,
    targetCohortTable = "cohort2",
    order = "first",
    targetCohortId = 1,
    window = c(1,Inf)
  )

cdm$cohort1NextEvent %>%
  glimpse()
```


Please note that the new columns added to the table (for all function presented) will have the format cohort_"cohort_definition_id"_ "time window". If the window number is negative, a "m" will be added in front to indicate it and no sign will be added to positive numbers.

## Example: addCohortIntersect function

`addCohortIntersect()`: Compute the intersect with a target cohort, you can compute the number of occurrences, a flag of presence, a certain date and/or the time difference.

We can use the function to compute all the intersect information with a target cohort. By default it will return output from `countCohortOccurrences()`, `dateOfCohort()`, `flagCohortPresence()` and `timeToCohort()` in a time frame. Use this function if you want to append all the intersection information with this function. For information on what these function does, you can read above example.

```{r, message= FALSE, warning=FALSE}
cdm$cohort1CohortIntersect <- cdm$cohort1 %>%
  addCohortIntersect(cdm = cdm,
    targetCohortTable = "cohort2",
    order = "first",
    targetCohortId = 1,
    window = c(1,Inf)
  )
cdm$cohort1CohortIntersect %>%
  glimpse()
```

We can also control which columns to append to your data by using the flag, count, date and time options in the function, if we do not want everything. For example if we only want the cohort count and flag we can use below example.

```{r, message= FALSE, warning=FALSE}
cdm$cohort1CohortIntersect <- cdm$cohort1 %>%
  addCohortIntersect(cdm = cdm,
    targetCohortTable = "cohort2",
    order = "first",
    targetCohortId = 1,
    window = c(1,Inf),
    flag = TRUE,
    count = TRUE,
    date = FALSE,
    time = FALSE
  )
cdm$cohort1CohortIntersect %>%
  glimpse()
```


