---
title: "Get cohort intersect using PatientProfiles"
author: "Marti Catala, Mike Du, Yuchen Guo and Kim Lopez-Guell"
output:
  html_document:
  pandoc_args: [
    "--number-offset=1,0"
  ]
number_sections: yes
toc: yes
vignette: >
 %\VignetteIndexEntry{Get cohort intersect using PatientProfiles}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(PatientProfiles)
```

## Introduction

This document demonstrate how to use the `addCohortIntersect()` function in PatientProfiles to obtain cohort intersect information for each individual in your CDM cohort table.

When using `addCohortIntersect()` you would need two input table.

1. Table containing cohort the individual for which the intersect information to be attached as extra columns. In the function this table is label as `x`. This table need to be saved as a table within CDM environment and contain.

2. Table where you search and compute the intersect information from. In the function this table is label as `cohortTableName`. This table need to be saved as a table within CDM environment.

Both table need to contain the column "cohort_definition_id", "subject_id", "cohort_start_date", "cohort_end_date". The intersect information this function can return are:

1. `number` count of the intersect in `cohortTableName`

2. `binary` indicator of whether intersect is found in `cohortTableName`

3. `date` the earliest or the latest date of the cohort_start_date of the intersect.

4. `time` the time in number days between the cohort_start_date of the individual and its earliest or the latest intersect date.

Below is an example of `addCohortIntersect()` generated using mock data.
```{r, message= FALSE, warning=FALSE, eval=FALSE}
library(DBI)
library(duckdb)
library(tibble)
library(PatientProfiles)
#functionality
cohort1 <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 1, 1, 1),
  subject_id = c(1, 1, 1, 2, 2),
  cohort_start_date = as.Date(
    c(
      "2020-01-01",
      "2020-01-15",
      "2020-01-20",
      "2020-01-01",
      "2020-02-01"
    )
  ),
  cohort_end_date = as.Date(
    c(
      "2020-01-01",
      "2020-01-15",
      "2020-01-20",
      "2020-01-01",
      "2020-02-01"
    )
  )
)
cohort2 <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 1, 1, 1, 1, 1),
  subject_id = c(1, 1, 1, 2, 2, 2, 1),
  cohort_start_date = as.Date(
    c(
      "2020-01-15",
      "2020-01-25",
      "2020-01-26",
      "2020-01-29",
      "2020-03-15",
      "2020-01-24",
      "2020-02-16"
    )
  ),
  cohort_end_date = as.Date(
    c(
      "2020-01-15",
      "2020-01-25",
      "2020-01-26",
      "2020-01-29",
      "2020-03-15",
      "2020-01-24",
      "2020-02-16"
    )
  ),
)
cdm <- mockPatientProfiles(cohort1=cohort1, cohort2=cohort2)
```

First we use `mockCohorProfiles()` to generate two cohort table in the cdm environment and name in `cohort1` and `cohort2` and saved it as cdm. Then to add the intersect information as column in `cohort1` from `cohort2`, we run below code.

```{r, message= FALSE, warning=FALSE, eval=FALSE}
cdm$cohort1 %>% addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "date") %>% 
  dplyr::rename(date_intersect = all_0_to_Inf) %>%
  addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "time") %>% 
    dplyr::rename(time_intersect = all_0_to_Inf) %>%
  addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "count") %>% 
    dplyr::rename(count_intersect = all_0_to_Inf) %>%
  addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "flag") %>%
  dplyr::rename(flag_intersect = all_0_to_Inf) 
```

As you see from the result above, it added 4 extra columns named "all_0_to_Inf" (they would have been by default "all_0_to_Inf", "all_0_to_Inf_1","all_0_to_Inf_2","all_0_to_Inf_3") before we renamed them to "date_intersect", "time_intersect", "count_intersect", "flag_intersect". The columns are name as "{cohortNames}_{window}_{window of interests}_{number_if_repeated}".

To change it to return to last cohort start date we can use the `order` option in the function

```{r, message= FALSE, warning=FALSE, eval=FALSE}
cdm$cohort1 %>% addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "date", order = "last")
```
The `value` option can be use to specified which value you want to return.

```{r, message= FALSE, warning=FALSE, eval=FALSE}
cdm$cohort1 %>% addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = c("flag"))
```

You can use the `window` option to change the look back period from date of reference in table x to date of event at event table.

```{r, message= FALSE, warning=FALSE, eval=FALSE}
cdm$cohort1 %>% addCohortIntersect(cdm = cdm,cohortTableName = "cohort2", value = "count", window = list(c(0,0)))
```
