---
title: "Plot Characteristics or large scale characteristics"
date: "`r Sys.Date()`"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Get cohort patients characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Different plots can be generated by function in this package. Here are some examples

##Large scale characteristics plot. First we create a mock data for the plot
```{r, eval=TRUE}
library(PatientProfiles)

person <- dplyr::tibble(
  person_id = c(1, 2),
  gender_concept_id = c(8507, 8532),
  year_of_birth = c(1990, 1992),
  month_of_birth = c(1, 1),
  day_of_birth = c(1, 1),
  race_concept_id = 0,
  ethnicity_concept_id = 0
)
observation_period <- dplyr::tibble(
  observation_period_id = c(1, 2),
  person_id = c(1, 2),
  observation_period_start_date = as.Date(c("2011-10-07", "2000-01-01")),
  observation_period_end_date = as.Date(c("2031-10-07", "2030-01-01")),
  period_type_concept_id = 44814724
)
cohort_interest <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 1, 2),
  subject_id = c(1, 1, 2, 2),
  cohort_start_date = as.Date(c(
    "2012-10-10", "2015-01-01", "2013-10-10", "2015-01-01"
  )),
  cohort_end_date = as.Date(c(
    "2012-10-10", "2015-01-01", "2013-10-10", "2015-01-01"
  ))
)
drug_exposure <- dplyr::tibble(
  drug_exposure_id = 1:11,
  person_id = c(rep(1, 8), rep(2, 3)),
  drug_concept_id = c(
    rep(1125315, 2), rep(1503328, 5), 1516978, 1125315, 1503328, 1516978
  ),
  drug_exposure_start_date = as.Date(c(
    "2010-10-01", "2012-12-31", "2010-01-01", "2012-09-01", "2013-04-01",
    "2014-10-31", "2015-05-01", "2015-10-01", "2012-01-01", "2012-10-01",
    "2014-10-12"
  )),
  drug_exposure_end_date = as.Date(c(
    "2010-12-01", "2013-05-12", "2011-01-01", "2012-10-01", "2013-05-01",
    "2014-12-31", "2015-05-02", "2016-10-01", "2012-01-01", "2012-10-30",
    "2015-01-10"
  )),
  drug_type_concept_id = 38000177,
  quantity = 1
)
condition_occurrence <- dplyr::tibble(
  condition_occurrence_id = 1:8,
  person_id = c(rep(1, 4), rep(2, 4)),
  condition_concept_id = c(
    317009, 378253, 378253, 4266367, 317009, 317009, 378253, 4266367
  ),
  condition_start_date = as.Date(c(
    "2012-10-01", "2012-01-01", "2014-01-01", "2010-01-01", "2015-02-01",
    "2012-01-01", "2013-10-01", "2014-10-10"
  )),
  condition_end_date = as.Date(c(
    "2013-01-01", "2012-04-01", "2014-10-12", "2015-01-01", "2015-03-01",
    "2012-04-01", "2013-12-01", NA
  )),
  condition_type_concept_id = 32020
)
cdm <- mockPatientProfiles(
  person = person, observation_period = observation_period,
  cohort_interest = cohort_interest, drug_exposure = drug_exposure,
  condition_occurrence = condition_occurrence
)

concept <- dplyr::tibble(
  concept_id = c(1125315, 1503328, 1516978, 317009, 378253, 4266367),
  domain_id = NA_character_,
  vocabulary_id = NA_character_,
  concept_class_id = NA_character_,
  concept_code = NA_character_,
  valid_start_date = as.Date("1900-01-01"),
  valid_end_date = as.Date("2099-01-01")
) %>%
  dplyr::mutate(concept_name = paste0("concept: ", .data$concept_id))

cdm <- CDMConnector::insertTable(cdm, "concept", concept)

largeScaleCharacteristicsResult <- cdm$cohort_interest %>%
  addDemographics(
    ageGroup = list(c(0, 24), c(25, 150))
  ) %>%
  summariseLargeScaleCharacteristics(
    strata = list("age_group", c("age_group", "sex")),
    episodeInWindow = c("condition_occurrence", "drug_exposure"),
    minimumFrequency = 0
  )
```

Use `plotLargeScaleCharacteristics()` to generate a scatterplot for large scale characteristics. User can specify x-axis (xAxis), y-axis (yAxis), variables to facet by (facetVars), variables to color by (colorVars), the order of facets (facetOrder), and whether to display the strings on x-axis vertically, as some concept id strings may be too long. It's worth noted if providing facetOrder, used need to make sure: 1. facetVars is specified 2. all values in facetOrder are in the unique combinations of facetVars. Below is an example.
```{r, eval=TRUE}
  levels_ordered <- c( "-365 to -31", "-30 to -1", "0 to 0", "1 to 30", "31 to 365")

  plotLargeScaleCharacteristics(
    data =  largeScaleCharacteristicsResult %>% dplyr::filter(group_level  ==  "cohort_1"),
    xAxis = "variable_name",
    yAxis = "estimate_value",
    facetVars = c("variable_level"),
    colorVars = c("group_level", "strata_level", "strata_name"),
    facetOrder = levels_ordered,
    vertical_x = TRUE
  )
  
```


## Characteristics plots
`plotCharacteristics` can plot the summariseCharacteristics results, with boxplot or barplot. First we create a mock data.

```{r, eval=TRUE}
observation_period <- dplyr::tibble(
  observation_period_id = c(1, 2, 3),
  person_id = c(1, 2, 3),
  observation_period_start_date = as.Date(c(
    "1985-01-01", "1989-04-29", "1974-12-03"
  )),
  observation_period_end_date = as.Date(c(
    "2011-03-04", "2022-03-14", "2023-07-10"
  )),
  period_type_concept_id = 0
)
dus_cohort <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 1, 2),
  subject_id = c(1, 1, 2, 3),
  cohort_start_date = as.Date(c(
    "1990-04-19", "1991-04-19", "2010-11-14", "2000-05-25"
  )),
  cohort_end_date = as.Date(c(
    "1990-04-19", "1991-04-19", "2010-11-14", "2000-05-25"
  ))
)
comorbidities <- dplyr::tibble(
  cohort_definition_id = c(1, 2, 2, 1),
  subject_id = c(1, 1, 3, 3),
  cohort_start_date = as.Date(c(
    "1990-01-01", "1990-06-01", "2000-01-01", "2000-06-01"
  )),
  cohort_end_date = as.Date(c(
    "1990-01-01", "1990-06-01", "2000-01-01", "2000-06-01"
  ))
)
medication <- dplyr::tibble(
  cohort_definition_id = c(1, 1, 2, 1),
  subject_id = c(1, 1, 2, 3),
  cohort_start_date = as.Date(c(
    "1990-02-01", "1990-08-01", "2009-01-01", "1995-06-01"
  )),
  cohort_end_date = as.Date(c(
    "1990-02-01", "1990-08-01", "2009-01-01", "1995-06-01"
  ))
)
emptyCohort <- dplyr::tibble(
  cohort_definition_id = numeric(),
  subject_id = numeric(),
  cohort_start_date = date(),
  cohort_end_date = date()
)
cdm <- mockPatientProfiles(
  dus_cohort = dus_cohort, cohort1 = emptyCohort,
  cohort2 = emptyCohort, observation_period = observation_period,
  comorbidities = comorbidities, medication = medication
)

cdm$dus_cohort <- omopgenerics::newCohortTable(
  table = cdm$dus_cohort, cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2), cohort_name = c("exposed", "unexposed")
  )
)
cdm$comorbidities <- omopgenerics::newCohortTable(
  table = cdm$comorbidities, cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2), cohort_name = c("covid", "headache")
  )
)
cdm$medication <- omopgenerics::newCohortTable(
  table = cdm$medication,
  cohortSetRef = dplyr::tibble(
    cohort_definition_id = c(1, 2, 3),
    cohort_name = c("acetaminophen", "ibuprophen", "naloxone")
  ),
  cohortAttritionRef = NULL
)
characteristicsResult <- summariseCharacteristics(
  cdm$dus_cohort,
  cohortIntersect = list(
    "Medications" = list(
      targetCohortTable = "medication", value = "flag", window = c(-365, 0)
    ), "Comorbidities" = list(
      targetCohortTable = "comorbidities", value = "flag", window = c(-Inf, 0)
    )
  )
)
```

Now we show barplot example by setting plotStyle = "barplot" in `plotCharacteristics`. Similar to previous function, user can define axes, facetVars and colorVars. Currently, barplot is only plotting percentage.
```{r, eval=TRUE}
plotCharacteristics(
    data =  characteristicsResult,
    xAxis = "estimate_value",
    yAxis = "variable_name",
    plotStyle = "barplot",
    facetVars = c("group_level"),
    colorVars = c("variable_name", "variable_level")
  )

```

User can plot boxplot based on q25 q75 median min max in data, using `plotCharacteristics` and set plotStyle = "boxplot".
```{r,eval=TRUE}
plotCharacteristics(
    data =  characteristicsResult,
    xAxis = "estimate_value",
    yAxis = "variable_name",
    plotStyle = "boxplot",
    facetVars = "variable_name",
    colorVars = c("group_level")
  )
```
