---
title: "Summarise result"
output: 
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{summarise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In previous vignettes we have seen how to add patient level demographics (age, sex, prior observation, ...) `r vignette("demographics", package = "PatientProfiles")` or intersections with cohorts `r vignette("cohort-intersect", package = "PatientProfiles")`, concepts and tables. 

Once we have added several columns to our table of interest we may want to summarise all this data into a `summarised_result` object using several different estimates.

We support different types of variables, variable type is assigned using `dplyr::type_sum`:

- Date: `date` or `dttm`.

- Numeric: `dbl` or `drtn`.

- Integer: `int` or `int64`.

- Categorical: `chr`, `fct` or `ord`.

- Logical: `lgl`.

Lets get started creating a our data that we are going to summarise:

```{r setup}
library(PatientProfiles)


```
